---
title: "Assignment API: Retrieving 1990 Census Data via IPUMS"
format: html
editor: visual
---

## Background

The U.S. Census Bureau's 1990 API endpoint is currently offline. To access 1990 Decennial data, we must pivot from `tidycensus` to the **IPUMS NHGIS** API using the `{ipumsr}` package.


## Part 1: Authentication

**Task:** Register at [nhgis.org](https://uma.pop.umn.edu/nhgis/user/new), generate an API key at [account.ipums.org](https://account.ipums.org), and set it in your R environment.

```{r}
#| label: setup
#| message: false
library(ipumsr)
library(dplyr)

# Set your API key (run this once)
# set_ipums_api_key("PASRE_YOUR_KEY_HERE", save = TRUE)
```




## Part 2: The Data Challenge

Your goal is to extract 1990 State Population Status (Table NP1) data for all States.




- To figure out which tables and variables you need, you use the [NHGIS Data Finder](https://data2.nhgis.org/) on the web. The API codes (like NH4 or NP1) correspond directly to the labels and categories you select in that interface. 

- To find the data from the [NHGIS Data Finder](https://data2.nhgis.org/). Log into your account,
then Filter Geographic level by STATE, YEARS by 1990 and TOPICS total population, the select the DATA, TABLE NAME, NP1, YEAR-DATASET, 1990_STF1



### Essential Functions Reference

To complete this, you will need to utilize these key functions from the `{ipumsr}` package:

-   `define_extract_agg()`: Defines which dataset, table, and geography you want.

-   `submit_extract()`: Sends your request to the IPUMS servers.

-   `wait_for_extract()`: Pauses R until the IPUMS server finishes building your file.

-   `download_extract()`: Downloads the completed ZIP file to your computer.

-   `read_nhgis()`: Loads the downloaded data into a data frame.

```{r}
#| label: extract-data

# 1. Define
# 2. Submit
# 3. Wait & Download
# 4. Read

```

## Part 3: Data Cleaning 

Using the metadata discovered via ``ipums_var_info()`.

Rename and the select variables

- gisjoin = GISJOIN,

- year = YEAR,

-  state_abbr = STUSAB,

-   division = DIVISIONA,

-   region  = REGIONA,

-   state_name = STATE,

-   state_code = STATEA,

-   land_area = AREALAND,

-   water_area = AREAWAT,

-   total_pop = ET1001


# Part 4: Find the population density
 population density = total population/ land area 
 
 
# Part 5: Find Water Share

water share = water area /(land area + water area)


### Tips for Students

-   **The Wait:** Remind them that `wait_for_extract()` is a "blocking" function; R will show a "busy" icon while IPUMS processes the request.

-   **The Metadata:** They should use the [NHGIS Data Finder](https://www.nhgis.org) to verify that Table **NP1** belongs to the **1990_STF1** (Summary Tape File 1) dataset.
